Optimizing the power of adaptive seamless designs using Bayesian Optimization

### 2020-12-09 main.tex
reading 2

Here, we propose to use a formal approach to optimization of the clinical trial design, in the situation of a very large number of available options for the parameters specifying the design. 
This set of candidate designs can be regarded as a multi-dimensional search space, including potentially many real-valued and categorical parameters.
The vast space of possible designs makes the optimization computationally demanding.
In the following, we demonstrate that the so-called Model-based optimization (MBO) framework~\citep{jones_taxonomy_2001} can be successfully applied to select a clinical trial design.
As evaluation metric we consider the specific case that the statistical power of the clinical trial design should be maximized, for given effect sizes and number of treatments.
Further, it is crucial that such a design is found in a short time.
%\subsection{Related work}
%subsub Clinical 
% TODO: Tim

MBO, often also called Bayesian optimization, has been successfully applied to optimize expensive black-boxes in many different scenarios.
Popular applications include hyper-parameter tuning of machine learning methods~\citep{snoek_practical_2012}, general algorithm configuration~\citep{hutter_sequential_2011}, and many more.
Various implementations of MBO methods are used widely, such as SMAC~\citep{hutter_sequential_2011}, Spearmint~\citep{snoek_practical_2012}, and BoTorch~\citep{balandat_botorch_2020}.
For this work we use the implementation in the R-package \texttt{mlrMBO}~\citep{bischl_mlrmbo_2017}.
This implementation has been successfully used to optimize hyper-parameters of machine learning methods on various tasks~\citep{bischl_mlrmbo_2017, wozniak_candle_2018}, and in the biomedical context it has been applied to optimize model-weights~\citep{richter_modelbased_2019,browaeys_nichenet_2020}.
Also, MBO is well suited for multi-criteria optimization.
Different adaptions of MBO exist~\citep{horn_modelbased_2015} that return a set of non-dominated points.
Many of them are implemented in \texttt{mlrMBO} as well.
However, in this work we restrict ourselves to single-criteria optimization.

The remainder of the manuscript is structured as follows.
In the next section we present as motivating example an application of adaptive seamless designs for chronic obstructive pulmonary disease (COPD).
In the sections~\ref{sec:adaptive_seamless_designs} and~\ref{sec:model_based_optimization}, we explain the background on adaptive seamless designs (asd) and model-based optimization (MBO), respectively.
In section~\ref{sec:simulation_study}, we present a simulation study that is closely related to the motivating example and in which the suitability of MBO for optimizing trial designs is demonstrated.
In the last Section~\ref{sec:discussion}, we discuss promising extensions as well as limitations of the approach.

@@

The general idea of this paper is to apply MBO for finding a trial design that is optimal with respect to a certain metric.
As an example, we consider the maximization of the statistical power of a trail design, given a set of parameters to be chosen for the trial design.
We compare the computation time as well as the maximal obtained power between the results obtained with MBO and with an exhaustive grid search over the space of adjustable parameters.
The presented optimization approach can also be applied for optimizing other metrics than the power, as long as the metric itself is numeric and the parameters of the trial design are numeric and include none or just few categorical choices.

One specific trial design that falls into this category are \emph{adaptive seamless designs}~\citep{barnes_integrating_2010} as implemented in the \texttt{asd}~\cite{parsons_software_2011} package and described in Section~\ref{sec:adaptive_seamless_designs}.
%(What is THE asd paper? peter bauer, meinter kieser, 1999 statistics in medicine, % https://doi.org/10.1002/(SICI)1097-0258(19990730)18:14%3C1833::AID-SIM221%3E3.0.CO;2-3, ) 
% https://doi.org/10.1002/bimj.200510232
% seamless trials (assoziert)
%  https://doi.org/10.1002/bimj.200510231
In~\citet{friede_adaptive_2020} the authors present an application of \emph{adaptive seamless designs} for chronic obstructive pulmonary disease (COPD).
We will use this COPD study to apply MBO for finding a trial design with maximal statistical power, given the parameters of the respective design options.
%within the given scenarios restrictions.

@@

\subsection{Two-stage design with treatment selection}
\label{ssec:two_stage_design_with_treatment_selection}
Here we consider a particular type of adaptive seamless designs, namely two-stage designs with treatment (or more specifically dose) selection. The trial starts in stage 1 with $K$ experimental treatments or doses of one experimental treatment which are compared to a common control. Interest is in assessing the treatment effects $\beta_k$ with $k=1,\dots, K$ comparing experimental treatment $k$ vs. control and testing the null hypotheses $H_{0,k}: \beta_k=0$ against the alternative  $H_{1,k}: \beta_k>0$ (larger $\beta_k$ being better). For the purpose of treatments selection, test statistics $Z_{1,k}$ based on data from the first design stage are considered with larger $Z_{1,k}$ indicating more beneficial treatments. These can use data on the primary outcome used for testing or any other outcome that is available at the time of the interim analysis. In practical applications, early outcomes are often used to inform interim decisions, since the primary endpoint might take a longer time to observe. In the interim analysis, treatments are selected for continuation into the second stage; the set of selected treatments is denoted by $\mathcal{S}$. The selection rules used here are described in Section \ref{ssec:selection_rules}. In the final analysis, the closed test principle is applied to account for the multiple hypotheses. The intersection hypotheses are tested by combining stagewise p-values for the respective intersection hypotheses using a prespecified combination function. Here we use the inverse normal combination function~\citep{lehmacher_adaptive_1999}
%FIXME Show function

@@

\subsection{Simulating ASD}
\label{ssec:simulating_asd}
R package asd by Nick Parsons
Test statistics are simulated, not individual participant data

@@

\begin{figure}[ht]
\centering
\begin{tikzpicture}[auto,node distance=1cm,align=center,text width = 3cm, inner sep = 0.1cm, font=\sffamily \scriptsize, outer sep= 0cm]
  \node[draw, ellipse, minimum height=1cm, text width=2.5cm] (ClinicalScenarios) at (0,0) {
    \textbf{Clinical scenarios}\\
    Simulation Studies};
  \node[name=block, rectangle split, rectangle split parts=2, draw, above = of ClinicalScenarios] (UnknownPoints)
  { \begin{itemize}[leftmargin=.3cm, topsep=0cm,itemsep=0cm,partopsep=0cm, parsep=0cm]
      \item Treatment effect on early interim outcome.
      \item Treatment effect on late disability outcome
    \end{itemize}
    \nodepart{second} Unknown
  };
  \node[name=block, rectangle split, rectangle split parts=2, draw, anchor = north east] (HistoricalDataPoints) at (UnknownPoints.north west)
  { \begin{itemize}[leftmargin=.3cm, topsep=0cm,itemsep=0cm,partopsep=0cm, parsep=0cm]
      \item Correlation between early interim and late disability outcomes.
    \end{itemize}
    \nodepart{second} Historical data
  };
  %\node[draw, inner xsep=1em, inner ysep=1em, above left = of ClinicalScenarios, fit = (HistoricalDataPoints) (UnknownPoints)] (boxDiseaseSpecific) {};
  \node[draw, text width = 6.2cm, anchor = south west] (DiseaseSpecific) at (HistoricalDataPoints.north west) {\textbf{Disease-specific features}};
  %\node[draw, above right = of ClinicalScenarios] (DesignOptions) {Design options};
  \node[name=block, rectangle split, rectangle split parts=2, draw, anchor = north west] (ConstrainedPoints) at ([xshift = 4cm]HistoricalDataPoints.north east)
  { \begin{itemize}[leftmargin=.3cm, topsep=0cm,itemsep=0cm,partopsep=0cm, parsep=0cm]
      \item Number of test treatments
      \item Total subjects in trial
      \item recruitment rate
    \end{itemize}
    \nodepart{second} Constrained by health care environment/infrastructure
  };
  \node[name=block, rectangle split, rectangle split parts=2, draw, anchor = north west] (CombinationPoints) at (ConstrainedPoints.north east)
  { \begin{itemize}[leftmargin=.3cm, topsep=0cm,itemsep=0cm,partopsep=0cm, parsep=0cm]
      \item Adaptive seamless design time point of interim analysis
      \item Treatment selection rules at interim Conventional phase 2/3
      \item split of resources between phases
      \item end oh phase futility criteria
    \end{itemize}
    \nodepart{second} Combination of options
  };
  \node[draw,text width = 6.2cm, anchor = south west] (DesignOptions) at (ConstrainedPoints.north west) {\textbf{Design Options}};
  \node[draw, below = 0.5cm of ClinicalScenarios] (DesignPerformance) {
    \textbf{Design performance} \\ e.g.\ statistical power};
  \node[draw, right = of DesignPerformance, text width = 4cm, node distance = 2cm] (Evaluation) {
    \textbf{Clinical Scenario Evaluation}\\
    Design performance measures evaluated across a wide range of clinical scenarios};
  \draw[->] (HistoricalDataPoints) -- (ClinicalScenarios);
  \draw[->] (UnknownPoints) -- (ClinicalScenarios);
  \draw[->] (CombinationPoints) -- (ClinicalScenarios);
  \draw[->] (ConstrainedPoints) -- (ClinicalScenarios);
  \draw[->] (ClinicalScenarios) -- (DesignPerformance);
  \draw[->] (DesignPerformance) -- (Evaluation);
\end{tikzpicture}
%Friede et al (2010) DIJ, Figure 2 https://link.springer.com/content/pdf/10.1177/009286151004400507.pdf? https://link.springer.com/content/pdf/10.1177/009286151004400607.pdf
\caption{Application of the refined clinical scenario evaluation framework to adaptive seamless designs in progressive multiple sclerosis.}
\label{fig:friede_refinement_2010_fig2}
\end{figure}

\section{Model-based Optimization}
\label{sec:model_based_optimization}

Sequential \textbf{Model-based Optimization (MBO)}~\citep{jones_taxonomy_2001} (also known as Bayesian optimization) is a state-of-the-art~\citep{shahriari_taking_2016} technique for expensive black-box optimization problems.
In comparison to other black-box optimization methods, like Genetic Algorithms or Simulated Annealing, MBO is especially suitable when evaluating a configuration (e.g.\ running a simulation with certain parameters, here denoted by $\theta$) is very time consuming.
Then it becomes infeasible to evaluate the black-box for thousands of configurations.

\subsection{General principle}
MBO solves an optimization problem within a bounded search space $\Theta$:
\[
\theta^\ast := \operatorname{arg\,max}_{\theta \in \Theta} f(\theta),
\]
where $f(\theta)$ denotes the evaluation of the black-box with the input configuration $\theta$.
To reduce the number of evaluations on $f$ the key idea of MBO is to only evaluate values of $\theta$ that are estimated to lead to a small value of $f(\theta)$.
The estimate is generated by a so called \emph{surrogate model}.
Typically, this is a regression model that predicts the outcome of $f$ based on previous evaluations of $f$.
First, an initial design of already evaluated configurations is needed.
Then, iteratively, the MBO algorithm fits the surrogate on the previous evaluations, proposes a new configuration $\theta$ and evaluates it on $f$.
The steps are repeated until a budget is exhausted.

@@

The goal of this simulation study is to investigate the potential benefit of MBO for effectively finding a clinical trial design that maximizes a metric.
The analysis is motivated by the clinical example with COPD patients introduced in Section~\ref{sec:motivating_example}, with treatment selection in the first of two stages.
In the first stage five arms (defined by the control and four different administered doses) are considered, and in the second stage only a selected subset of arms is kept for further evaluation. 
The metric to be optimized is the statistical power, where a positive outcome means a trial arm with true effect is detected as significant effect.

@@

In the simulation study we compare the ability to two optimization approaches for finding an optimal configuration, on the one hand an exhaustive grid search on the parameter space and on the other hand  MBO (Model-based optimization).
For different scenarios we compare the statistical power of the best found clinical trial design.
We also compare the runtime needed to determine the respective optimization results.
Next we introduce the two optimization strategies.

@@

\paragraph{MBO (Model-based optimization)}
For applying MBO, we first define how the surrogate model that predicts the outcome of $\tilde{f}$ for unknown values of $\theta$ (parameter sets) is generated.
As regression model for the surrogate we choose Kriging (also called Gaussian process regression) and use the implementation in the R-package \texttt{DiceKriging}~\citep{roustant_dicekriging_2012}, because it is known to work well for fairly low dimensional search spaces.
We configure Kriging to apply the Mattern$\frac{5}{2}$ kernel with an estimated \emph{nugget effect} to account for the noisy response of $f$ and without scaling the input variables to $[0,1]$.

@@

As an acquisition function we choose the $\operatorname{AEI}$ as explained in Section~\ref{sec:model_based_optimization}.
We start MBO with an initial design of $16$ randomly sampled points and allow $100$ further iterations, summing up to a total budget of $116$ evaluated configurations of the simulator $\tilde{f}$.

@@

The whole MBO procedure is repeated $20$ times to ensure that a single optimization run was not just lucky in finding the optimum.

Additionally we investigate how \emph{MBO} performs when it is limited to return configurations that are within the grid.
Therefore we introduce \textbf{MBO Grid}, which executes the optimization exactly as \emph{MBO}, but the best configuration is mapped to the closest value on the grid. 
In other words, for \emph{MBO Grid} we conduct the independent evaluation using the configuration within the grid that is closest to the optimal configuration that is proposed by \emph{MBO}.
The comparison of the results of \emph{MBO Gird} and \emph{Gird} will allow two possible conclusions:
First, if \emph{MBO Grid} is worse then \emph{MBO}, we can assume that the resolution of \emph{Grid} is not high enough to find an optimal configuration.
Second, if \emph{MBO Grid} is better then \emph{Grid}, it shows that \emph{Grid} did not select the optimal configuration, due to noise.
Summing up, there are two factors that can prevent \emph{Grid} from finding the optimal configuration: lack of resolution and noise of the outcome.

@@

The algorithms used in this study are implemented in R.
For the model-based optimization, the R-package mlrMBO~\citep{bischl_mlrmbo_2017} is used.
For the simulation of clinical trials, the function \texttt{treatsel.sim} of the R-package asd~\citep{parsons_software_2011} is used, with parameter values given in Table~\ref{tab:par_implement}.
\begin{table}[h]
  \caption{Parameter choices for the R function \texttt{treatsel.sim}}
  \label{tab:par_implement}
  \centering
  \begin{tabular}{lll}
  \hline
  Meaning of parameter & Parameter & Value \\
  \hline
  Number of simulation iterations & \texttt{nsim} & $1000$ \\
  Correlation between early and final outcomes & \texttt{corr} & $0.4$ \\
  Test level & \texttt{level} & $0.025$ \\
  Vector of treatment numbers for determining power & \texttt{ptest} & $(3,4)$ \\
  \hline
  \end{tabular}
\end{table}
The vector of treatment numbers for determining power \texttt{ptest} counts rejections of 3 or 4 hypotheses ???.
Note that the parameter \texttt{nsim} denotes the internal simulation iterations of $\tilde{f}$. 
The ten additional stochastic repetitions we conduct for the \emph{grid search} and \emph{MBO} are additional replicates.

@@

In this chapter we present the results of the four optimization strategies: exhaustive \emph{Grid} search, grid search with a reduced resolution (\emph{Grid Small}), \emph{MBO} and \emph{MBO Grid},
We focus our comparison on independent, replicated evaluations of the configurations proposed by each optimizer.
The replicated evaluations give us an unbiased estimate of the performance ($y_{\text{valid}}$) of each optimizer.

@@

The plots show that \emph{MBO} generally gives results that are comparable to those obtained with the (exhaustive) \emph{Grid} method, for all effects sets. 
Only in some cases \emph{Grid} has a slight advantage (\emph{effect: linear}, $n_{\text{treat}} = 2000$; \emph{effect: sigmoid}, $n_{\text{treat}} = 1000$; \emph{effect: sigmoid}, $n_{\text{treat}} = 2000$).
Below we will see that these scenarios have in common that the optimal configuration is close to the border of the search space. 
In other cases \emph{MBO} yields slightly better results (\emph{effect: paper}, $n_{\text{treat}} = 2000$; \emph{effect: paper2}, $n_{\text{treat}} = 500$; \emph{effect: sigmoid}, $n_{\text{treat}} = 500$).
Across all scenarios, \emph{MBO} and \emph{Grid} perform similar, even though \emph{MBO} due to its evaluation budget is allowed only 116 evaluations in contrast to the 1350 evaluations for \emph{Grid}, with a resolution of 25 points per dimension. 
The difference in the number of evaluations is clearly reflected in the average runtime of the optimization methods given in Table~\ref{tab:table_time}.
\emph{MBO} is approximately 20 times faster.
\input{generated/tables/table_time.tex}

For \emph{MBO Grid}, only solutions on the grid that was evaluated by \emph{Grid} are considered in the evaluation. 
Comparing the results of \emph{MBO Grid} to those of \emph{MBO} we mainly see similar performances for both.
This indicates, that the resolution of \emph{Grid} is fine enough to achieve performance similar to \emph{MBO}.

Comparing \emph{MBO Grid} to \emph{Gird} we can observe, that for the scenarios \emph{effect: paper}, $n_{\text{treat}} = 500$ and \emph{effect: paper}, $n_{\text{treat}} = 1000$ \emph{MBO Grid} achieves slightly better performance then \emph{Grid}.
The \emph{Grid} optimization selected the best outcome "optimistically" from all observed outcome and has no notion of the stochasticity of the problem, i.e. in this case the noise on the predicted power. 
Accordingly, we assume that for the mentioned scenarios the noise was so high, that \emph{Gird} was misguided by overoptimistic outcomes when selecting the final configuration as optimization result.
The independent validation $y_{\text{valid}}$ represents the unbiased performance of this final configuration, and the overoptimism becomes apparent.
In these scenarios \emph{MBO} performs better, since the final configuration is determined based on the mean prediction of the surrogate of \emph{MBO}, as explained in Section~\ref{ssec:best_point}.
Below we will study on a single scenario if a reduced noise due to a higher value of \texttt{nsim} decreases the effect of the overly optimistic selection of \emph{Grid}.

If \emph{MBO} was not able to find a better solution then \emph{Grid}, then also \emph{MBO Grid} cannot yield better results as it just takes the configuration on the Grid that is closest to the configuration found by \emph{MBO}.
An increase in performance of \emph{MBO Grid} compared to \emph{MBO} is purely due to chance.

>>>>

Here, we propose to use a formal approach to optimization of the clinical trial design, in the situation of a very large number of available options for the parameters specifying the design. 
This set of candidate designs can be regarded as a multi-dimensional search space, including potentially many real-valued and categorical parameters.
The vast space of possible designs makes the optimization computationally demanding.
In the following, we demonstrate that the so-called \emph{Bayesian optimization} (BO) framework~\citep{jones_taxonomy_2001} can be successfully applied to select a clinical trial design.
As evaluation metric we consider the specific case that the statistical power of the clinical trial design should be maximized, for given effect sizes and number of treatments.
Further, it is crucial that such a design is found in a short time.
%\subsection{Related work}
%subsub Clinical 
% TODO: Tim

BO, often also called Model-based optimization, has been successfully applied to optimize expensive black-boxes in many different scenarios.
Popular applications include hyper-parameter tuning of machine learning methods~\citep{snoek_practical_2012}, general algorithm configuration~\citep{hutter_sequential_2011}, and many more.
Various implementations of BO methods are used widely, such as SMAC~\citep{hutter_sequential_2011}, Spearmint~\citep{snoek_practical_2012}, and BoTorch~\citep{balandat_botorch_2020}.
For this work we use the implementation in the R-package \texttt{mlrMBO}~\citep{bischl_mlrmbo_2017}.
This implementation has been successfully used to optimize hyper-parameters of machine learning methods on various tasks~\citep{bischl_mlrmbo_2017, wozniak_candle_2018}, and in the biomedical context it has been applied to optimize model-weights~\citep{richter_modelbased_2019,browaeys_nichenet_2020}.
Also, BO is well suited for multi-criteria optimization.
Different adaptions of BO exist~\citep{horn_modelbased_2015} that return a set of non-dominated points.
Many of them are implemented in \texttt{mlrMBO} as well.
However, in this work we restrict ourselves to single-criteria optimization.

The remainder of the manuscript is structured as follows.
In the next section we present as motivating example an application of adaptive seamless designs for chronic obstructive pulmonary disease (COPD).
In the sections~\ref{sec:adaptive_seamless_designs} and~\ref{sec:bayesian_optimization}, we explain the background on adaptive seamless designs (asd) and Bayesian optimization (BO), respectively.
In section~\ref{sec:simulation_study}, we present a simulation study that is closely related to the motivating example and in which the suitability of BO for optimizing trial designs is demonstrated.
In the last Section~\ref{sec:discussion}, we discuss promising extensions as well as limitations of the approach.

@@

The general idea of this paper is to apply BO for finding a trial design that is optimal with respect to a certain metric.
As an example, we consider the maximization of the statistical power of a trail design, given a set of parameters to be chosen for the trial design.
We compare the computation time as well as the maximal obtained power between the results obtained with BO and with an exhaustive grid search over the space of adjustable parameters.
The presented optimization approach can also be applied for optimizing other metrics than the power, as long as the metric itself is numeric and the parameters of the trial design are numeric and include none or just few categorical choices.

One specific trial design that falls into this category are \emph{adaptive seamless designs}~\citep{barnes_integrating_2010} as implemented in the \texttt{asd}~\cite{parsons_software_2011} package and described in Section~\ref{sec:adaptive_seamless_designs}.
%(What is THE asd paper? peter bauer, meinter kieser, 1999 statistics in medicine, % https://doi.org/10.1002/(SICI)1097-0258(19990730)18:14%3C1833::AID-SIM221%3E3.0.CO;2-3, )  bauer_combining_1999
% https://doi.org/10.1002/bimj.200510232 bretz_confirmatory_2006 
% seamless trials (assoziert) 
%  https://doi.org/10.1002/bimj.200510231 schmidli_confirmatory_2006
In~\citet{friede_adaptive_2020} the authors present an application of \emph{adaptive seamless designs} for chronic obstructive pulmonary disease (COPD).
We will use this COPD study to apply BO for finding a trial design with maximal statistical power, given the parameters of the respective design options.
%within the given scenarios restrictions.

@@

\subsection{Two-stage design with treatment selection}
\label{ssec:two_stage_design_with_treatment_selection}
Here we consider a particular type of adaptive seamless designs, namely two-stage designs with treatment (or more specifically dose) selection. The trial starts in stage 1 with $K$ experimental treatments or doses of one experimental treatment which are compared to a common control. Interest is in assessing the treatment effects $\beta_k$ with $k=1,\dots, K$ comparing experimental treatment $k$ vs. control and testing the null hypotheses $H_{0,k}: \beta_k=0$ against the alternative  $H_{1,k}: \beta_k>0$ (larger $\beta_k$ being better). For the purpose of treatments selection, test statistics $Z_{1,k}$ based on data from the first design stage are considered with larger $Z_{1,k}$ indicating more beneficial treatments. These can use data on the primary outcome used for testing or any other outcome that is available at the time of the interim analysis. In practical applications, early outcomes are often used to inform interim decisions, since the primary endpoint might take a longer time to observe. In the interim analysis, treatments are selected for continuation into the second stage; the set of selected treatments is denoted by $\mathcal{S}$. The selection rules used here are described in Section \ref{ssec:selection_rules}. In the final analysis, the closed test principle is applied to account for the multiple hypotheses. The intersection hypotheses are tested by combining stagewise $p$-values for the respective intersection hypotheses using a prespecified combination function. Here we use the inverse normal combination function~\citep{lehmacher_adaptive_1999}:
\begin{equation}
  \frac{1}{\sqrt{K}} \sum_{k=1}^K \Phi^{-1}(1-p_k) \,
\end{equation}
where $\Phi^{-1}$ denotes the quantile function of the standard normal distribution.

@@

\subsection{Simulating ASD}
\label{ssec:simulating_asd}
The R-package \texttt{asd} by~\citet{parsons_software_2011} simulates the test statistics and not individual participant data.

@@

\begin{figure}[ht]
\centering
\begin{tikzpicture}[auto,node distance=1cm,align=center,text width = 3cm, inner sep = 0.1cm, font=\sffamily \scriptsize, outer sep= 0cm]
  \node[draw, ellipse, minimum height=1cm, text width=2.5cm] (ClinicalScenarios) at (0,0) {
    \textbf{Clinical scenarios}\\
    Simulation Studies};
  \node[name=block, rectangle split, rectangle split parts=2, draw, above = of ClinicalScenarios] (UnknownPoints)
  { \begin{itemize}[leftmargin=.3cm, topsep=0cm,itemsep=0cm,partopsep=0cm, parsep=0cm]
      \item Treatment effect on early interim outcome
      \item Treatment effect on late disability outcome
    \end{itemize}
    \nodepart{second} Unknown
  };
  \node[name=block, rectangle split, rectangle split parts=2, draw, anchor = north east] (HistoricalDataPoints) at (UnknownPoints.north west)
  { \begin{itemize}[leftmargin=.3cm, topsep=0cm,itemsep=0cm,partopsep=0cm, parsep=0cm]
      \item Correlation between early interim and late disability outcomes
    \end{itemize}
    \nodepart{second} Historical data
  };
  %\node[draw, inner xsep=1em, inner ysep=1em, above left = of ClinicalScenarios, fit = (HistoricalDataPoints) (UnknownPoints)] (boxDiseaseSpecific) {};
  \node[draw, text width = 6.2cm, anchor = south west] (DiseaseSpecific) at (HistoricalDataPoints.north west) {\textbf{Disease-specific features}};
  %\node[draw, above right = of ClinicalScenarios] (DesignOptions) {Design options};
  \node[name=block, rectangle split, rectangle split parts=2, draw, anchor = north west] (ConstrainedPoints) at ([xshift = 4cm]HistoricalDataPoints.north east)
  { \begin{itemize}[leftmargin=.3cm, topsep=0cm,itemsep=0cm,partopsep=0cm, parsep=0cm]
      \item Number of test treatments
      \item Total subjects in trial
      \item Recruitment rate
    \end{itemize}
    \nodepart{second} Constrained by health care environment/infrastructure
  };
  \node[name=block, rectangle split, rectangle split parts=2, draw, anchor = north west] (CombinationPoints) at (ConstrainedPoints.north east)
  { Adaptive seamless design
    \begin{itemize}[leftmargin=.3cm, topsep=0cm,itemsep=0cm,partopsep=0cm, parsep=0cm]
      \item time point of interim analysis
      \item treatment selection rules at interim Conventional phase 2/3
      \item split of resources between phases
      \item end oh phase futility criteria
    \end{itemize}
    \nodepart{second} Combination of options
  };
  \node[draw,text width = 6.2cm, anchor = south west] (DesignOptions) at (ConstrainedPoints.north west) {\textbf{Design Options}};
  \node[draw, below = 0.5cm of ClinicalScenarios] (DesignPerformance) {
    \textbf{Design performance} \\ e.g.\ statistical power};
  \node[draw, right = of DesignPerformance, text width = 4cm, node distance = 2cm] (Evaluation) {
    \textbf{Clinical Scenario Evaluation}\\
    Design performance measures evaluated across a wide range of clinical scenarios};
  \draw[->] (HistoricalDataPoints) -- (ClinicalScenarios);
  \draw[->] (UnknownPoints) -- (ClinicalScenarios);
  \draw[->] (CombinationPoints.223) -- (ClinicalScenarios);
  \draw[->] (ConstrainedPoints) -- (ClinicalScenarios);
  \draw[->] (ClinicalScenarios) -- (DesignPerformance);
  \draw[->] (DesignPerformance) -- (Evaluation);
\end{tikzpicture}
%Friede et al (2010) DIJ, Figure 2 https://link.springer.com/content/pdf/10.1177/009286151004400507.pdf? https://link.springer.com/content/pdf/10.1177/009286151004400607.pdf
\caption{Application of the refined clinical scenario evaluation framework to adaptive seamless designs in progressive multiple sclerosis.}
\label{fig:friede_refinement_2010_fig2}
\end{figure}

\section{Bayesian Optimization}
\label{sec:bayesian_optimization}

Bayesian Optimization (BO)~\citep{jones_taxonomy_2001}, also known as Model-based Optimization (MBO), is a state-of-the-art~\citep{shahriari_taking_2016} technique for expensive black-box optimization problems.
In comparison to other black-box optimization methods, like Genetic Algorithms or Simulated Annealing, BO is especially suitable when evaluating a configuration (e.g.\ running a simulation with certain parameters, here denoted by $\theta$) is very time consuming.
Then it becomes infeasible to evaluate the black-box for thousands of configurations.

\subsection{General principle}
BO solves an optimization problem within a bounded search space $\Theta$:
\[
\theta^\ast := \operatorname{arg\,max}_{\theta \in \Theta} f(\theta),
\]
where $f(\theta)$ denotes the evaluation of the black-box with the input configuration $\theta$.
To reduce the number of evaluations on $f$ the key idea of BO is to only evaluate values of $\theta$ that are estimated to lead to a small value of $f(\theta)$.
The estimate is generated by a so called \emph{surrogate model}.
Typically, this is a regression model that predicts the outcome of $f$ based on previous evaluations of $f$.
First, an initial design of already evaluated configurations is needed.
Then, iteratively, the BO algorithm fits the surrogate on the previous evaluations, proposes a new configuration $\theta$ and evaluates it on $f$.
The steps are repeated until a budget is exhausted.

@@

The goal of this simulation study is to investigate the potential benefit of BO for effectively finding a clinical trial design that maximizes a metric.
The analysis is motivated by the clinical example with COPD patients introduced in Section~\ref{sec:motivating_example}, with treatment selection in the first of two stages.
In the first stage five arms (defined by the control and four different administered doses) are considered, and in the second stage only a selected subset of arms is kept for further evaluation. 
The metric to be optimized is the statistical power, where a positive outcome means a trial arm with true effect is detected as significant effect.

@@

In the simulation study we compare the ability to two optimization approaches for finding an optimal configuration, on the one hand an exhaustive grid search on the parameter space and on the other hand BO (Bayesian optimization).
For different scenarios we compare the statistical power of the best found clinical trial design.
We also compare the runtime needed to determine the respective optimization results.
Next we introduce the two optimization strategies.

@@

\paragraph{BO (Model-based optimization)}
For applying BO, we first define how the surrogate model that predicts the outcome of $\tilde{f}$ for unknown values of $\theta$ (parameter sets) is generated.
As regression model for the surrogate we choose Kriging (also called Gaussian process regression) and use the implementation in the R-package \texttt{DiceKriging}~\citep{roustant_dicekriging_2012}, because it is known to work well for fairly low dimensional search spaces.
We configure Kriging to apply the Mattern$\frac{5}{2}$ kernel with an estimated \emph{nugget effect} to account for the noisy response of $f$ and without scaling the input variables to $[0,1]$.

@@

As an acquisition function we choose the $\operatorname{AEI}$ as explained in Section~\ref{sec:bayesian_optimization}.
We start BO with an initial design of $16$ randomly sampled points and allow $100$ further iterations, summing up to a total budget of $116$ evaluated configurations of the simulator $\tilde{f}$.

@@

The whole BO procedure is repeated $20$ times to ensure that a single optimization run was not just lucky in finding the optimum.

Additionally we investigate how \emph{BO} performs when it is limited to return configurations that are within the grid.
Therefore we introduce \textbf{BO Grid}, which executes the optimization exactly as \emph{BO}, but the best configuration is mapped to the closest value on the grid. 
In other words, for \emph{BO Grid} we conduct the independent evaluation using the configuration within the grid that is closest to the optimal configuration that is proposed by \emph{BO}.
The comparison of the results of \emph{BO Gird} and \emph{Gird} will allow two possible conclusions:
First, if \emph{BO Grid} is worse then \emph{BO}, we can assume that the resolution of \emph{Grid} is not high enough to find an optimal configuration.
Second, if \emph{BO Grid} is better then \emph{Grid}, it shows that \emph{Grid} did not select the optimal configuration, due to noise.
Summing up, there are two factors that can prevent \emph{Grid} from finding the optimal configuration: lack of resolution and noise of the outcome.

@@

The algorithms used in this study are implemented in R.
For the Bayesian optimization, the R-package \texttt{mlrMBO}~\citep{bischl_mlrmbo_2017} is used.
For the simulation of clinical trials, the function \texttt{treatsel.sim} of the R-package \texttt{asd}~\citep{parsons_software_2011} is used, with parameter values given in Table~\ref{tab:par_implement}.
\begin{table}[h]
  \caption{Parameter choices for the R function \texttt{treatsel.sim}}
  \label{tab:par_implement}
  \centering
  \begin{tabular}{lll}
  \hline
  Meaning of parameter & Parameter & Value \\
  \hline
  Number of simulation iterations & \texttt{nsim} & $1000$ \\
  Correlation between early and final outcomes & \texttt{corr} & $0.4$ \\
  Test level & \texttt{level} & $0.025$ \\
  Vector of treatment numbers for determining power & \texttt{ptest} & $(3,4)$ \\
  \hline
  \end{tabular}
\end{table}
The vector of treatment numbers for determining power \texttt{ptest} counts rejections of 3 or 4 hypotheses ???.
Note that the parameter \texttt{nsim} denotes the internal simulation iterations of $\tilde{f}$. 
The ten additional stochastic repetitions we conduct for the \emph{grid search} and \emph{BO} are additional replicates.

@@

In this chapter we present the results of the four optimization strategies: exhaustive \emph{Grid} search, grid search with a reduced resolution (\emph{Grid Small}), \emph{BO} and \emph{BO Grid},
We focus our comparison on independent, replicated evaluations of the configurations proposed by each optimizer.
The replicated evaluations give us an unbiased estimate of the performance ($y_{\text{valid}}$) of each optimizer.

@@

The plots show that \emph{BO} generally gives results that are comparable to those obtained with the (exhaustive) \emph{Grid} method, for all effects sets. 
Only in some cases \emph{Grid} has a slight advantage (\emph{effect: linear}, $n_{\text{treat}} = 2000$; \emph{effect: sigmoid}, $n_{\text{treat}} = 1000$; \emph{effect: sigmoid}, $n_{\text{treat}} = 2000$).
Below we will see that these scenarios have in common that the optimal configuration is close to the border of the search space. 
In other cases \emph{BO} yields slightly better results (\emph{effect: paper}, $n_{\text{treat}} = 2000$; \emph{effect: paper2}, $n_{\text{treat}} = 500$; \emph{effect: sigmoid}, $n_{\text{treat}} = 500$).
Across all scenarios, \emph{BO} and \emph{Grid} perform similar, even though \emph{BO} due to its evaluation budget is allowed only 116 evaluations in contrast to the 1350 evaluations for \emph{Grid}, with a resolution of 25 points per dimension. 
The difference in the number of evaluations is clearly reflected in the average runtime of the optimization methods given in Table~\ref{tab:table_time}.
\emph{BO} is approximately 20 times faster.
\input{generated/tables/table_time.tex}

For \emph{BO Grid}, only solutions on the grid that was evaluated by \emph{Grid} are considered in the evaluation. 
Comparing the results of \emph{BO Grid} to those of \emph{BO} we mainly see similar performances for both.
This indicates, that the resolution of \emph{Grid} is fine enough to achieve performance similar to \emph{BO}.

Comparing \emph{BO Grid} to \emph{Gird} we can observe, that for the scenarios \emph{effect: paper}, $n_{\text{treat}} = 500$ and \emph{effect: paper}, $n_{\text{treat}} = 1000$ \emph{BO Grid} achieves slightly better performance then \emph{Grid}.
The \emph{Grid} optimization selected the best outcome "optimistically" from all observed outcome and has no notion of the stochasticity of the problem, i.e. in this case the noise on the predicted power. 
Accordingly, we assume that for the mentioned scenarios the noise was so high, that \emph{Gird} was misguided by overoptimistic outcomes when selecting the final configuration as optimization result.
The independent validation $y_{\text{valid}}$ represents the unbiased performance of this final configuration, and the overoptimism becomes apparent.
In these scenarios \emph{BO} performs better, since the final configuration is determined based on the mean prediction of the surrogate of \emph{BO}, as explained in Section~\ref{ssec:best_point}.
Below we will study on a single scenario if a reduced noise due to a higher value of \texttt{nsim} decreases the effect of the overly optimistic selection of \emph{Grid}.

If \emph{BO} was not able to find a better solution then \emph{Grid}, then also \emph{BO Grid} cannot yield better results as it just takes the configuration on the Grid that is closest to the configuration found by \emph{BO}.
An increase in performance of \emph{BO Grid} compared to \emph{BO} is purely due to chance.
